Nail: A practical interface generator for data formats

### 2014-04-02 paper/impl.tex
make spell

A generated Nail parser makes two passes through the input: the first to
validate and recognise the input, and the second to bind this data to the internal
model. Currently the parser is a straightforward top-down parser, although
facilities have been made to add Packrat parsing to achieve linear parsing
times.

@@

Under this premise, the actions need limited information from the
recogniser to correctly handle the input stream. In particular, the parser's
control flow branches only at the choice, repetition, and constant combinators.
Thus, for each of those combinators, we store the minimum amount of information
required to reconstruct the syntactic structure of the input. The trace is an
array of integers.
Whenever the parser encounters a choice, it appends two integers to the trace. After it
successfully parses a choice, the parser writes the number of that choice and
the length of the trace when it began parsing that choice. When backtracking in
the input, the parser does not backtrack in the trace. This means that offsets
within the trace can be used for a Packrat hash-table to memoize backtrack-heavy
parsers.

@@

To demonstrate the feasibility of the Nail parser generator, our parser
generator uses a Nail parser to recognise Nail grammars. A superset of the
grammar language described in this paper is implemented in 100 lines of Nail,
which feed into about 1,000 lines of C++ that implement the actual parser
generator. Bootstrapping is supported via an implementation of the Nail language
in the Lemon parser generator, a variant of Yacc.

>>>>

A generated Nail parser makes two passes through the input: the first to
validate and recognize the input, and the second to bind this data to the internal
model. Currently the parser is a straightforward top-down parser, although
facilities have been made to add Packrat parsing to achieve linear parsing
times.

@@

Under this premise, the actions need limited information from the
recognizer to correctly handle the input stream. In particular, the parser's
control flow branches only at the choice, repetition, and constant combinators.
Thus, for each of those combinators, we store the minimum amount of information
required to reconstruct the syntactic structure of the input. The trace is an
array of integers.
Whenever the parser encounters a choice, it appends two integers to the trace. After it
successfully parses a choice, the parser writes the number of that choice and
the length of the trace when it began parsing that choice. When backtracking in
the input, the parser does not backtrack in the trace. This means that offsets
within the trace can be used for a Packrat hash-table to memoize backtrack-heavy
parsers.

@@

To demonstrate the feasibility of the Nail parser generator, our parser
generator uses a Nail parser to recognize Nail grammars. A superset of the
grammar language described in this paper is implemented in 100 lines of Nail,
which feed into about 1,000 lines of C++ that implement the actual parser
generator. Bootstrapping is supported via an implementation of the Nail language
in the Lemon parser generator, a variant of Yacc.

### 2014-04-02 paper/relwk.tex
make spell

A subclass of logic errors are so-called \textit{weird machines}, where
implementation side effects or under-specified parser behaviour leads to a
protocol or data format inadvertently becoming a Turing-complete execution
environment, even though the original grammar did not require it.
Frequently, this execution environment can either
then directly manipulate data in unwanted ways or be used to make
exploiting another bug feasible.\footnote{For example, by compiling a
return-oriented-programming exploit from code fragments discovered on
the fly.} Examples include x86 page tables~\cite{bangert2013page}, and
ELF symbols and relocations~\cite{shapiro2013weird}. In the offensive
research community, this has been generalised into treating a program
as a \textit{weird machine}~\cite{bratus2011exploit} that operates on
an input, analogous to a virtual machine operating on bytecode.
Nail avoids these problems by having the parser precisely match the
specified grammar, eliminating under-specified behavior.

@@

Parser generators for binary protocols were first introduced by the
Hammer~\cite{hammer-parser} parser.
While previous parser generators could also be used to write grammars
for binary protocols,\footnote{Theoretically speaking, the alphabet
over which a grammar is an  abstract set, so most algorithms work
just as well on an alphabet of $\{0,1\}$.} doing so is practically
inconvenient. Hammer allows the programmer to specify a grammar in
terms of bits and bytes instead of characters. Common concerns, such as
endianness and bit-packing are transparently hidden by the library. Hammer
implements grammars as language-integrated parser combinators, an approach
popularised by Parsec for Haskell~\cite{LeijenMeijer:parsec}. The
parser combinator style (to our knowledge, first  described
in~\cite{burge1975recursive}) is a natural way of concisely expressing
top-down grammars~\cite{Danielsson:2010:TPC:1863543.1863585}\footnote{For
more background on the history of expressing grammars, see Bryan Ford's
masters thesis~\cite{ford2002packrat}, which also describes the default
parsing algorithm used by Hammer.} by composing them from one or multiple
sub-parsers.  Hammer then constructs a tree of function pointers which
can be invoked to parse a given input into an abstract syntax tree (AST).
\XXX[how does Nail relate to Hammer, in terms of what's new, what it
solves, etc?]

@@

Unfortunately,  parser frameworks are seldom used  to recognise
machine-created input. For example, the security-critical and
well-engineered MIT Kerberos distribution uses parser generators, but
only for handling configuration files. A notable exception is the Mongrel
web server\footnote{\url{http://mongrel2.org/}} which uses a grammar
for HTTP written in the Ragel~\cite{ragel-paper} regular expression
language. Mongrel was re-written from scratch multiple times to achieve
better scaleability and design, yet the grammar could be  re-used across
all iterations~\cite{patterson-citation}.

>>>>

A subclass of logic errors are so-called \textit{weird machines}, where
implementation side effects or under-specified parser behavior leads to a
protocol or data format inadvertently becoming a Turing-complete execution
environment, even though the original grammar did not require it.
Frequently, this execution environment can either
then directly manipulate data in unwanted ways or be used to make
exploiting another bug feasible.\footnote{For example, by compiling a
return-oriented-programming exploit from code fragments discovered on
the fly.} Examples include x86 page tables~\cite{bangert2013page}, and
ELF symbols and relocations~\cite{shapiro2013weird}. In the offensive
research community, this has been generalized into treating a program
as a \textit{weird machine}~\cite{bratus2011exploit} that operates on
an input, analogous to a virtual machine operating on bytecode.
Nail avoids these problems by having the parser precisely match the
specified grammar, eliminating under-specified behavior.

@@

Parser generators for binary protocols were first introduced by the
Hammer~\cite{hammer-parser} parser.
While previous parser generators could also be used to write grammars
for binary protocols,\footnote{Theoretically speaking, the alphabet
over which a grammar is an  abstract set, so most algorithms work
just as well on an alphabet of $\{0,1\}$.} doing so is practically
inconvenient. Hammer allows the programmer to specify a grammar in
terms of bits and bytes instead of characters. Common concerns, such as
endianness and bit-packing are transparently hidden by the library. Hammer
implements grammars as language-integrated parser combinators, an approach
popularized by Parsec for Haskell~\cite{LeijenMeijer:parsec}. The
parser combinator style (to our knowledge, first  described
in~\cite{burge1975recursive}) is a natural way of concisely expressing
top-down grammars~\cite{Danielsson:2010:TPC:1863543.1863585}\footnote{For
more background on the history of expressing grammars, see Bryan Ford's
masters thesis~\cite{ford2002packrat}, which also describes the default
parsing algorithm used by Hammer.} by composing them from one or multiple
sub-parsers.  Hammer then constructs a tree of function pointers which
can be invoked to parse a given input into an abstract syntax tree (AST).
\XXX[how does Nail relate to Hammer, in terms of what's new, what it
solves, etc?]

@@

Unfortunately,  parser frameworks are seldom used  to recognize
machine-created input. For example, the security-critical and
well-engineered MIT Kerberos distribution uses parser generators, but
only for handling configuration files. A notable exception is the Mongrel
web server\footnote{\url{http://mongrel2.org/}} which uses a grammar
for HTTP written in the Ragel~\cite{ragel-paper} regular expression
language. Mongrel was re-written from scratch multiple times to achieve
better scalability and design, yet the grammar could be  re-used across
all iterations~\cite{patterson-citation}.
